### 2019-06-09 2-Chapters/6-Chapter/chapter6.tex
Done for proofreading Ch6
Done proof reading ch5

In this last chapter, we study a generalization of the stationary multi-armed bandit model,
to account for possible non-stationarity of the rewards.
% % that is, for problems that could be tackled with MAB algorithms.
% Instead of dealing with changes happening possibly at every time step, like in adversarial models, we rather consider a finite small number $\Upsilon_T$ of \emph{change-points}, and we consider problems that are stationary on any interval between two consecutive change-points.
%
We review existing works on piece-wise stationary MAB models.
% for which the two main families of proposed solutions are either passively or actively adaptive. In practice, it was already observed that actively adaptive strategies usually greatly outperform the passively adaptive ones.
% The more efficient strategies from the current state-of-the-art are based on a combination of an efficient algorithm for the stationary problem with an efficient online change detection algorithm, like CUSUM, PHT or the Generalized Likelihood Ratio Test (GLRT), and so we pursue on this direction.
%
We study the Generalized Likelihood Ratio Test (GLRT) for
% Using finite-time results of a recent paper studying the GLRT for sub-Gaussian distributions, we then extend the results to the GLRT for any 
bounded or sub-Bernoulli distributions.
Our test enjoys finite-time guarantees
% its \emph{false alarm probability}, meaning that in a stationary model, the test should not detect any change,
% and its \emph{detection delay}, meaning that in a model with a change-point, the test should detect within a reasonable delay.
%
of its \emph{false alarm probability} and \emph{detection delay}, and can be combined with an efficient bandit policy (\klUCB), to propose an efficient algorithm for piece-wise stationary problems, \GLRklUCB.
% Our policy takes ideas from the two most recent and most efficient policies, \CUSUMUCB{} and \MUCB, that enjoys good regret bounds of the order of $R_T = \bigO{\sqrt{\Upsilon_T T \log(T)}}$, if $T$ is the horizon and $\Upsilon_T$ the number of change-points (known before hand).
% % \MUCB{} restarts the underlying \UCB{} bandit policy on all arms when a change is detected, while \CUSUMUCB{} restarts the observations of only the arm on which a change is detected.
%
% We give the first unified analysis of a piece-wise stationary bandit algorithm for the two variants, and our analysis
We analyze its regret, and show that it achieves state-of-the-art non-asymptotic regret bounds.
% and yield regret bounds with explicit constants.
% Moreover, our regret bounds are the first ones to be of the order $\bigO{\sqrt{\Upsilon_T T \log(T)}}$, when the algorithm runs with no prior knowledge of the problem other than the horizon $T$ and the number of change-points $\Upsilon_T$ (both \MUCB{} and \CUSUMUCB{} require to know before hand a certain measure $\Delta^{\text{change}}$ of the difficulty of the problem).
%
Finally, we compare our approach against other state-of-the-art solutions on extensive numerical experiments.
% , on different piece-wise stationary bandit problems.
% % , of increasing difficulty.
% %
% We also include additional details about our policy \GLRklUCB, including a sensitivity analysis of its two parameters, a numerical evaluation of its time and memory costs, as well as details on an easy but efficient way to speed up the B-GLRT test while not loosing too much in terms of regrets.

\TODOL{C'est bizarre ce problème d'espacement vertical ici... je ne sais pas comment le corriger !}

>>>>

In this last chapter, we study a generalization of the stationary multi-armed bandit model,
to account for possible non-stationarity of the rewards.
% % that is, for problems that could be tackled with MAB algorithms.
% Instead of dealing with changes happening possibly at every time step, like in adversarial models, we rather consider a finite small number $\Upsilon_T$ of \emph{change-points}, and we consider problems that are stationary on any interval between two consecutive change-points.
%
We review existing works on piece-wise stationary MAB models.
% for which the two main families of proposed solutions are either passively or actively adaptive. In practice, it was already observed that actively adaptive strategies usually greatly outperform the passively adaptive ones.
% The more efficient strategies from the current state-of-the-art are based on a combination of an efficient algorithm for the stationary problem with an efficient online change detection algorithm, like CUSUM, PHT or the Generalized Likelihood Ratio Test (GLRT), and so we pursue on this direction.
%
We study the Generalized Likelihood Ratio Test (GLRT) for
% Using finite-time results of a recent paper studying the GLRT for sub-Gaussian distributions, we then extend the results to the GLRT for any 
bounded or sub-Bernoulli distributions.
Our test enjoys finite-time guarantees
% its \emph{false alarm probability}, meaning that in a stationary model, the test should not detect any change,
% and its \emph{detection delay}, meaning that in a model with a change-point, the test should detect within a reasonable delay.
%
of its \emph{false alarm probability} and \emph{detection delay}, and can be combined with an efficient bandit policy (\klUCB), to propose an efficient algorithm for piece-wise stationary problems, \GLRklUCB.
% Our policy takes ideas from the two most recent and most efficient policies, \CUSUMUCB{} and \MUCB, that enjoys good regret bounds of the order of $R_T = \bigO{\sqrt{\Upsilon_T T \log(T)}}$, if $T$ is the horizon and $\Upsilon_T$ the number of change-points (known before hand).
% % \MUCB{} restarts the underlying \UCB{} bandit policy on all arms when a change is detected, while \CUSUMUCB{} restarts the observations of only the arm on which a change is detected.
%
% We give the first unified analysis of a piece-wise stationary bandit algorithm for the two variants, and our analysis
We analyze its regret, and show that it achieves state-of-the-art non-asymptotic regret bounds.
% and yield regret bounds with explicit constants.
% Moreover, our regret bounds are the first ones to be of the order $\bigO{\sqrt{\Upsilon_T T \log(T)}}$, when the algorithm runs with no prior knowledge of the problem other than the horizon $T$ and the number of change-points $\Upsilon_T$ (both \MUCB{} and \CUSUMUCB{} require to know before hand a certain measure $\Delta^{\text{change}}$ of the difficulty of the problem).
%
Finally, we compare our approach with other state-of-the-art solutions on large numerical experiments.
% , on different piece-wise stationary bandit problems.
% % , of increasing difficulty.
% %
% We also include additional details about our policy \GLRklUCB, including a sensitivity analysis of its two parameters, a numerical evaluation of its time and memory costs, as well as details on an easy but efficient way to speed up the B-GLRT test while not loosing too much in terms of regrets.

% \TODOL{C'est bizarre ce problème d'espacement vertical ici... je ne sais pas comment le corriger !}

### 2019-06-09 2-Chapters/6-Chapter/nonstatbandits.git/NonStatB.tex
Lots of small typos.
Done proof reading ch5

\textbf{Outline.}
%
% This chapter is structured as follows.
We introduce the model and review related works in Section~\ref{sec:6:BanditSetting}. In Section~\ref{sec:6:ChangePointDetector}, we study the Generalized Likelihood Ratio test for sub-Bernoulli distributions (B-GLRT) as a Change-Point Detector (CPD) algorithm.
We introduce the two variants of \GLRklUCB{} algorithm in Section~\ref{sec:6:GLRklUCB_Algorithm}, where we also present upper bounds on the regret of both variants.
The unified regret analysis for these two algorithms is given in Section~\ref{sec:6:RegretAnalysis}.
Numerical experiments are presented in Section~\ref{sec:6:NumericalExperiments}, with more details in the Appendix.

@@

A \emph{piece-wise stationary bandit model} is characterized by a set of $K$ arms.
A (random) stream of rewards $(Y_{k,t})_{t\in\N^*}$ is associated to each arm $k \in [K]$. We assume that the rewards are bounded, and without loss of generality we assume that $Y_{k,t} \in [0,1]$.
We denote by $\mu_k(t) \eqdef  \bE[Y_{k,t}]$ the mean reward of arm $k$ at round $t$. At each round $t$, a decision maker has to select an arm $A(t)\in[K]$, based on past observation and receives the corresponding reward $r(t) = Y_{A(t),t}$.
At time $t$, we denote by $k_t^*$ an arm with maximal expected reward, \ie, $\mu_{k_t^*}(t) = \max_k \mu_k(t)$, called an optimal arm (possibly not unique).

A policy $\cA$ chooses the next arm to play based on the sequence of past plays and obtained rewards.
Like for stationary problems (see Definition~\ref{def:2:regret}), the performance of $\cA$ is measured by its (piece-wise stationary) \emph{regret}, the difference between the expected reward obtained by an oracle policy playing an optimal arm $k_t^*$ at time $t$, and that of the policy $\cA$
We use the same notation as the regret for stationary problem without ambiguity, as in this chapter we only consider the following definition.
%
\begin{equation}\label{def:6:nti}
    R_T^{\cA} = \E\left[\sum_{t=1}^T \left(\mu_{k_t^*}(t) - \mu_{A(t)}(t)\right)\right].
\end{equation}

In the piece-wise \iid{} model, we furthermore assume that there is a (relatively small) number of \emph{breakpoints}, denoted by $\Upsilon_T \eqdef \sum_{t=1}^{T-1} \indic\left(\exists k\in[K]: \mu_k(t) \neq \mu_k(t+1)\right)$.
If we denote $\tau^{(0)} = 0$, we define the $i$-th breakpoint recursively by
$\tau^{(i)} = \inf\{t > \tau^{(i-1)} : \exists k : \mu_k(t) \neq \mu_k(t+1)\}$.
%
Hence for $t\in[\tau^{(i)} + 1,\tau^{(i+1)}]$, that is on any stationary segments, the rewards $(Y_{k,t})$ associated to each arm $k$ are \iid{} (hence the name of \emph{piece-wise stationary} problems).

Note than when a breakpoint occurs, we do not assume that all the arms means  change, but that \emph{there exists} an arm whose mean has changed.
Depending on the application, many scenario can be meaningful: changes occurring on all arms simultaneously (due to some exogenous event), or only one or a few arms change at each breakpoint.
For instance, for a cognitive radio application in a IoT-like network, we can imagine that all the objects of one company use a fixed channel or subset of channels, and on a particular day in a particular city, if the company deploys a lot of new objects, then one or some channels see their mean occupancy rates abruptly change.

@@

\textbf{An interesting interpolation.}
%
The piece-wise stationary bandit model can be viewed as an interpolation between stationary and adversarial models, as the stationary model corresponds to $\Upsilon_T = 0$, while some adversarial models can be considered as a special (worst) case, with $\Upsilon_T = T$ (when considering an adaptive or an oblivious adversary).
However, analyzing an algorithm for the piece-wise stationary bandit model requires to assume a small number of changes, typically $\Upsilon_T = o(\sqrt{T})$.
Note that the adversarial model of \cite{Auer02NonStochastic} is quite powerful, and the authors presented in Section~8 of their paper the EXP3.S algorithm for the piece-wise stationary problem, as explained in Section~\ref{sec:6:Introduction} above.

@@

We review previous works that studied the piece-wise stationary bandit model, or variants of this model.
To our knowledge, all the previous works are based on the idea of combining a classical bandit policy, such as Thompson sampling, \UCB{} or EXP3, with a strategy to account for changes in arms' distributions.
Following the vocabulary used in previous works, we make the distinction between passively and actively adaptive strategies.
Active strategies monitors the rewards of each arm by using a change detection algorithm \cite{Basseville93}, and reset the history of pulls and rewards of one or all the arms as soon as a change is detected.
Passive strategies use a (fixed) discount factor or a limited memory size, while most active strategies use a growing memory.
% , while active strategies are usually tuned with a threshold for the statistical test.

@@

\begin{corollary}\label{cor:6:Local}
\begin{leftbar}[corollarybar]  % XXX leftbar corollarybar, comment if needed
    For ``easy'' problems satisfying the corresponding Assumption~\ref{ass:6:LongPeriods},
    with $\Delta^{\text{opt}}$ and ${\Delta}^{\text{change}}$ defined as in Corollary~\ref{cor:6:Global}, the following holds.
    \begin{enumerate}
        \item Choosing $\alpha = \sqrt{\ln(T) / T}$, $\delta = 1 / \sqrt{T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K}{\left({\Delta}^{\text{change}}\right)^2} \EffChange\sqrt{T\ln(T)} + 1 / \left(\Delta^{\text{opt}}\right)^2} \EffChange\ln(T)},
        \end{equation}
        \item Choosing $\alpha = \sqrt{\Upsilon_T\ln(T) / T}$, $\delta = 1 / \sqrt{\Upsilon_T T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K^2}{\left({\Delta}^{\text{change}}\right)^2}\sqrt{\Upsilon_T T\ln(T)} + \frac{K\Upsilon_T}{\left(\Delta^{\text{opt}}\right)^2}\ln(T)},
        \end{equation}
        \item Choosing $\alpha = \sqrt{\EffChange\ln(T) / T}$, $\delta = 1 / \sqrt{\EffChange T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K}{\left({\Delta}^{\text{change}}\right)^2}\sqrt{\EffChange T\ln(T)} + \frac{\EffChange}{\left(\Delta^{\text{opt}}\right)^2}\ln(T)},
        \end{equation}
        \item Choosing $\alpha = \sqrt{K\Upsilon_T\ln(T) / T}$, $\delta = 1 / \sqrt{K\Upsilon_T T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K}{\left({\Delta}^{\text{change}}\right)^2}\sqrt{\EffChange T\ln(T)} + \frac{1}{\left(\Delta^{\text{opt}}\right)^2} \EffChange\ln(T)}.
        \end{equation}
    \end{enumerate}
\end{leftbar}  % XXX leftbar corollarybar, comment if needed
\end{corollary}

@@

% -----------------------------------------------------------------
\begin{figure}[h!]  % [htbp]
    \centering
    \includegraphics[width=1.09\linewidth]{2-Chapters/6-Chapter/nonstatbandits.git/figures/SP__K3_T5000_N1000__9_algos_pb1/main_HistogramsRegret_shareX____env1-1_1316297932259102962.pdf}
    \caption{Histograms of the distributions of regret $R_T$ ($T=5000$) for problem 1.}
    \label{fig:6:histogramRegretPb1}
\end{figure}

>>>>

\textbf{Outline.}
%
% This chapter is structured as follows.
We introduce the model in Section~\ref{sec:6:BanditSetting}, and review related works in Section~\ref{sec:6:relatedWork}.
In Section~\ref{sec:6:ChangePointDetector} we study the Generalized Likelihood Ratio test for sub-Bernoulli distributions (B-GLRT) as a Change-Point Detector (CPD) algorithm.
We introduce the two variants of \GLRklUCB{} algorithm in Section~\ref{sec:6:GLRklUCB_Algorithm}, where we also present upper bounds on the regret of both variants.
The unified regret analysis is given in Section~\ref{sec:6:RegretAnalysis}, we prove one result in Section~\ref{sec:6:proofRegret}.
We discuss numerical experiments in Section~\ref{sec:6:NumericalExperiments}, with more details in the Appendix.

@@

A \emph{piece-wise stationary bandit model} is characterized by a set of $K$ arms.
A (random) stream of rewards $(Y_{k,t})_{t\in\N^*}$ is associated to each arm $k \in [K]$. We assume that the rewards are bounded, and without loss of generality we assume that $Y_{k,t} \in [0,1]$.
We denote by $\mu_k(t) \eqdef  \bE[Y_{k,t}]$ the mean reward of arm $k$ at round $t$. At each round $t$, a decision maker has to select an arm $A(t)\in[K]$, based on past observation and receives the corresponding reward $r(t) = Y_{A(t),t}$.
At time $t$, we denote by $k_t^*$ an arm with maximal expected reward, \ie, $\mu_{k_t^*}(t) \eqdef \max_k \mu_k(t)$, called an optimal arm (possibly not unique).

A policy $\cA$ chooses the next arm to play based on the sequence of past plays and obtained rewards.
Like for stationary problems (see Definition~\ref{def:2:regret} in Chapter~\ref{chapter:2}), the performance of $\cA$ is measured by its (piece-wise stationary) \emph{regret}, the difference between the expected reward obtained by an oracle policy playing an optimal arm $k_t^*$ at time $t$, and that of the policy $\cA$
We use the same notation as the regret for stationary problem without ambiguity, as in this chapter we only consider the following definition,
%
\begin{equation}\label{def:6:nti}
    R_T^{\cA} \eqdef \E\left[\sum_{t=1}^T \left(\mu_{k_t^*}(t) - \mu_{A(t)}(t)\right)\right].
\end{equation}

In the piece-wise \iid{} model, we furthermore assume that there is a (relatively small) number of \emph{breakpoints}, denoted by $\Upsilon_T \eqdef \sum_{t=1}^{T-1} \indic\left(\exists k\in[K]: \mu_k(t) \neq \mu_k(t+1)\right)$.
If we denote $\tau^{(0)} \eqdef 0$, we define the $i$-th breakpoint recursively by
$\tau^{(i)} \eqdef \inf\{t > \tau^{(i-1)} : \exists k : \mu_k(t) \neq \mu_k(t+1)\}$.
%
Hence for $t\in[\tau^{(i)} + 1,\tau^{(i+1)}]$, that is on any stationary segments, the rewards $(Y_{k,t})$ associated to each arm $k$ are \iid{} (hence the name of \emph{piece-wise stationary} problems).

Note than when a breakpoint occurs, we do not assume that all the arms means  change, but that \emph{there exists} an arm whose mean has changed.
Depending on the application, many scenario can be meaningful: changes occurring on all arms simultaneously (due to some exogenous event), or only one or a few arms change at each breakpoint.
For instance, for a cognitive radio application in a IoT-like network, we can imagine that all the objects of one company use a fixed (or a subset of) channel(s), and on a particular day in a city, if the company deploys a lot of new objects, then one (or some) channel(s) see their mean occupancy rates abruptly change.

@@

\textbf{An interesting interpolation.}
%
The piece-wise stationary bandit model can be viewed as an interpolation between stationary and adversarial models, as the stationary model corresponds to $\Upsilon_T = 0$, while some adversarial models can be considered as a special (worst) case, with $\Upsilon_T = T-1$ (when considering an adaptive or an oblivious adversary).
However, analyzing an algorithm for the piece-wise stationary bandit model requires to assume a small number of changes, typically $\Upsilon_T = o(\sqrt{T})$.
Note that the adversarial model of \cite{Auer02NonStochastic} is quite powerful, and the authors presented in Section~8 of their paper the EXP3.S algorithm for the piece-wise stationary problem, as explained in Section~\ref{sec:6:Introduction} above.

@@

We review previous works that studied the piece-wise stationary bandit model, or variants of this model.
To our knowledge, all the previous works are based on the idea of combining a classical bandit policy, such as Thompson sampling, \UCB{} or EXP3, with a strategy to account for changes in arms' distributions.
Following the vocabulary used in previous works, we make the distinction between passively and actively adaptive strategies.
Active strategies monitor the arms' rewards, by using change detection algorithms \cite{Basseville93}, and reset the history of pulls and rewards of one or all the arms as soon as a change is detected.
Passive strategies use a (fixed) discount factor or a limited memory size, while most active strategies use a growing memory.
% , while active strategies are usually tuned with a threshold for the statistical test.

@@

\begin{corollary}\label{cor:6:Local}
\begin{leftbar}[corollarybar]  % XXX leftbar corollarybar, comment if needed
    For ``easy'' problems satisfying the corresponding Assumption~\ref{ass:6:LongPeriods},
    with $\Delta^{\text{opt}}$ and ${\Delta}^{\text{change}}$ defined as in Corollary~\ref{cor:6:Global}, the following holds.
    \begin{enumerate}
        \item Choosing $\alpha = \sqrt{\ln(T) / T}$, $\delta = 1 / \sqrt{T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K}{\left(\Delta^{\text{change}}\right)^2} \EffChange\sqrt{T\ln(T)} + \frac{1}{\left(\Delta^{\text{opt}}\right)^2} \EffChange\ln(T)},
        \end{equation}
        \item Choosing $\alpha = \sqrt{\Upsilon_T\ln(T) / T}$, $\delta = 1 / \sqrt{\Upsilon_T T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K^2}{\left({\Delta}^{\text{change}}\right)^2}\sqrt{\Upsilon_T T\ln(T)} + \frac{K\Upsilon_T}{\left(\Delta^{\text{opt}}\right)^2}\ln(T)},
        \end{equation}
        \item Choosing $\alpha = \sqrt{\EffChange\ln(T) / T}$, $\delta = 1 / \sqrt{\EffChange T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K}{\left({\Delta}^{\text{change}}\right)^2}\sqrt{\EffChange T\ln(T)} + \frac{\EffChange}{\left(\Delta^{\text{opt}}\right)^2}\ln(T)},
        \end{equation}
        \item Choosing $\alpha = \sqrt{K\Upsilon_T\ln(T) / T}$, $\delta = 1 / \sqrt{K\Upsilon_T T}$ gives
        \begin{equation}
            R_T = \bigO{\frac{K}{\left({\Delta}^{\text{change}}\right)^2}\sqrt{\EffChange T\ln(T)} + \frac{1}{\left(\Delta^{\text{opt}}\right)^2} \EffChange\ln(T)}.
        \end{equation}
    \end{enumerate}
\end{leftbar}  % XXX leftbar corollarybar, comment if needed
\end{corollary}

@@

% -----------------------------------------------------------------
\begin{figure}[b!]  % [htbp]
    \centering
    \includegraphics[width=1.09\linewidth]{2-Chapters/6-Chapter/nonstatbandits.git/figures/SP__K3_T5000_N1000__9_algos_pb1/main_HistogramsRegret_shareX____env1-1_1316297932259102962.pdf}
    \caption{Histograms of the distributions of regret $R_T$ ($T=5000$) for problem 1.}
    \label{fig:6:histogramRegretPb1}
\end{figure}
